Competing Constructions for German Measure Phrases

### 2017-03-24 Latex/article.tex
Work on interpretation

%\subsection{Tools for modelling alternations}
%\label{sec:righttools}
%
%In this section, I discuss the relevant statistical approaches to modelling alternations based on corpus data.
%I mainly focus on well-established Generalised Linear Models (GLMs) and their \textit{multi-level} or \textit{hierarchical} extensions, which are also called \textit{mixed models} (GLMMs).
%
%% TODO CONTENT HERE
%
%In Section~\ref{sec:corpusnonhierarchicalmodel}, I briefly discuss why a non-hierarchical model -- while it is an option in principle -- has serious disadvantages for the data set under discussion (and similar data sets).
%In Section~\ref{sec:corpushierarchicalmodel}, I present the hierarchical model which is actually interpreted as the primary result of this study.
%Section~\ref{sec:bayesian} takes a short detour and shows (by example) that results from Bayesian estimation are not intrinsically different from those obtained from Maximum Likelihood methods, although Bayesian approaches have recently been touted to solve various problems facing applied statistics in various fields, even including cognitively oriented corpus linguistics.
%Finally, I summarise and interpret the results of the corpus study in Section~\ref{sec:modelssummary}.

@@

%Related to the , dative measure nouns favor the PGC more than nominative and accusative (also \citealp{Hentschel1993,Zimmer2015}).
%As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (PGC) would be preferred over \textit{mit einem Stück frischem Brot} (NAC).
%I propose
%

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always (but not exclusively) occur in the \PGCa\ and virtually never in the \NACa\ (see Section~\ref{sec:descriptive}).
More indirectly, however, plurality of the measure noun could also favor the \PGCa\ over the \NACa, and exact quantification in the form of cardinals even more so.%
\footnote{This preference was observed empirically by \cite{Hentschel1993}.}
I provide a highly illustrative example using a measure noun which has virtually developed two variants, one of which grammaticalised to a much higher degree, namely \textit{Haufen} `heap\slash lot'.
The phrase \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (\NACa) has essentially two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former use, however, is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps of French cotton' (\PGCa) -- not `two lots of French cotton'.
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of discrete portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.
The statistical models in Section~\ref{sec:corpusstudies} will be designed to detect such a subtle effect, if it exists.
%\footnote{In the extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
%Also note that the presentation of the examples in NAC and PGC is suggestive.
%Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
%However, it was made clear from the beginning and throughout that the preferences are indeed subtle.}

@@

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (unequivocal) PGC and NAC are positively correlated with the relative frequency the PGC and the NAC are chosen in the alternating case.
  \item MNPs with masculine dative kind nouns disfavour the GPC less.
  \item Classes of more clearly grammaticalised measure nouns favor the NAC.
  \item Plural measure nouns and measure nouns quantified by cardinals favor the PGC variant.
  \item The NAC is associated more strongly with higher registers.
\end{enumerate}

@@

\subsection{Annotation}
\label{sec:annotation}

@@

To capture the influence of register or style mentioned in Section~\ref{sec:analyses}, several proxy variables were used.
At the document level, the DECOW corpus has an annotation for \textit{Badness}.
As described in \cite{SchaeferEa2013}, \textit{Badness} measures how well the distribution of highly frequent short words in the document matches a pre-generated language model for German.
Documents with higher Badness usually contain more incoherent language, shorter sentences, etc.
If the PGC actually favors higher registers and styles, a high \textit{Badness} should be correlated with fewer occurrences of the PGC.
Documents in DECOW14 have also been annotated with a variable called \textit{Genitives}.
The higher the values of this variable, the lower the proportion of genitives among all case-bearing forms is.
While more genitives are also indicative of higher registers, the use of this variable as a regressor in the present study might be considered problematic.
Since the PGC contains a genitive itself, the regressor variable \textit{Genitive} and the document-level variable \textit{Genitives} are not fully independent.
However, since the PGCs make up for only a minute fraction of all genitives, I still use \textit{Genitives} as a regressor with the appropriate caveats.
Since word length is also a good indicator of higher registers and styles, the length of the match in characters (for example 17 for \textit{Kilo tolle Schoki} `a kilo of nice choc' and 49 for \textit{ein Kilogramm vorzüglicher Edelbitter-Schokolade} ` a kilogram of excellent fine dark chocolate') was added as another proxy to register.%
\footnote{The new version of the corpus contains over 100 document-level variables in the spirit of \cite{Biber1988}.
Unfortunately, these annotations were not available when my samples were taken, and even cruder proxies to register than Biberian features (like the length of the query match) had to be used.
All in all, all variables used as indicators of register here have to be interpreted with extreme caution.
}
Furthermore, abbreviated measure nouns like \textit{kg} or \textit{m} might be indicative of less elaborate technical styles, and hence the variable \textit{Measureabbreviated} was included, which is 1 if the measure noun is abbreviated.

The variables \textit{Measurecase} and \textit{Measurenumber} simply encode the grammatical case and number of the measure noun.
The variable \textit{Minus1pos} encodes the part-of-speech of the word preceding the measure noun.
The level names stand for \textit{adjective}, \textit{cardinal}, \textit{determiner}, \textit{other}, and \textit{preposition} (in the same order as in Table~\ref{tab:variables}).
The purpose of this variable is mainly to see whether cardinals really favor the \PGCa\ as hypothesised in Section~\ref{sec:analyses}.

@@

Last, to control for lemma frequency effects, the logarithm-transformed frequency per 1,000,000 words of each lemma was added from the frequency lists distributed by the DECOW designers (variables \textit{Kindfreq}, \textit{Measurefreq}).
All numeric variables were standardised and centered (z-transformed) to facilitate their interpretation in the regression models reported 
in the next section.

@@

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation, and \textit{rstanarm} \citep{rstanarm} for Markov-Chain Monte Carlo estimation.
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that either the \NACa\ or the \PGCa\ is chosen.
As explained in Section~\ref{sec:descriptive}, due to extensive case syncrestism in feminine NPs, feminine kind nouns in MNPs like \textit{(mit) einem Liter kalter Milch} `with a litre of cold milk' can either be datives (case agreement, \NACa) or genitives (\PGCa) whenever the measure noun is in the dative.
All dative MNPs with feminine kind nouns had to be removed, and the regressor \textit{Measurecase} consequently only had two levels when the kind noun was feminine.%
\footnote{Together with other cleanups (erroneous hits, extremely noisy material), the removal of such uninformative sentences reduced the total size of the sample to 5,063 sentences.}
Therefore, two separate models for masculine\slash neuter and feminine kind nouns were fitted.
For the masculine\slash neuter model, the final sample size was $n_{\text{MN}}=3,949$ with 672 cases of \PGCa\ and 3,277 cases of \NACa.
For the feminine model, it was $n_{\text{F}}=1,114$ with 462 cases of \PGCa\ and 652 cases of $\NACa$.
All regressors from Table~\ref{tab:variables} were included to begin with, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The regressors listed in Table~\ref{tab:variables} with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \cite[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.

@@

Regressors were removed if no convergence could be achieved (even after tuning the optimizer) or perfect separation occurred.
Beyond that, no step-down model selection was performed (see Section~\ref{sec:rightstatistics}).
Generalized variance inflation factors for the remaining regressors were calculated to check for multicollinearity \citep{FoxMonette1992}.
Except for the levels of lemma noun class factors \textit{Kindconsistency} and \textit{Measureclass}, virtually no regressor has a GVIF over 3.
Exceptions were GVIF(\textit{Kindattraction})=3.17 and GVIF(\textit{Kindfreq})=3.92, both still being far from dramatic.
The high GVIFs for the noun lemma classes (up to 46.2) correlate with a low degree of theoretical motivation (at least of the concrete classes as annotated).
However, re-estimating the models with these variables removed did \textit{not} alter the sign or magnitude of any of the other regressors, and I left them in the model.

@@

Standard model diagnostics show that the models perform very well.
For the masculine\slash neuter model, Nakagawa \& Schielzeth's determination coefficient is $R_m^2=0.495$ and $R^2_c=0.544$ (see \citealp{Gries2015} for a linguist-friendly introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.872$, which means an increase over the baseline of $\lambda=0.247$.
For the feminine model, $R_m^2=0.468$ and $R^2_c=0.574$ are achieved, as well as a correct prediction rate of $0.815$, which means an improvement over the baseline of $\lambda=0.554$.

In Table~\ref{tab:bigtable}, there are groups of columns for the results obtained from Maximum Likelihood estimation and Markov-Chain Monte Carlo estimation, both for the masculine\slash neuter and the feminine models.
In the remainder of this section, I describe how these results were obtained.
The coefficient estimates are specified for each regressor (level) in the columns labelled \textit{Coef}.%
\footnote{Empty cells in the table correspond to regressors and their levels which had to be removed because of convergence problems.}
For a robust estimation of the precision of the estimation, I ran a parametric bootstrap (using the \textit{confint.merMod} function from \textit{lme4}) with 1,000 replications and the bootstrap percentile method.
The resulting 95\% bootstrap intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{Low}, \textit{High} (for the upper and lower 2.5th percentile) and \textit{Width} (for the width of the interval).
The column \textit{Not 0} shows an asterisk for those intervals that do not include 0 as a more reasonable replacement of `significance levels'.

As mentioned above, not step-down model selection was performed.
However, for each regressor, a $p$ value was obtained by dropping it from the full model, re-eastimating the nested model and comparing it to the full model.
Instead of inexact Wald approximations or Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test with the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}.
These bootstrapped $p$ values are given in the columns labelled \textit{p\Sub{BP}} in Table~\ref{tab:bigtable}.

Finally, MCMC TODO.

%\begin{figure}[h]
%\centering
%\includegraphics[width=0.8\textwidth]{figures/corpus/04_glmm_fixef_firstlevel}
%\caption{Estimated first-level effects in GLMMs for Masculine\slash Neuter and Feminine}
%\label{fig:glmmfixef}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.8\textwidth]{figures/corpus/04_glmm_fixef_secondlevel_kind}
%\caption{Estimated second-level effects for kind lemmas in GLMMs for Masculine\slash Neuter and Feminine}
%\label{fig:glmmfixef}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.8\textwidth]{figures/corpus/04_glmm_fixef_secondlevel_measure}
%\caption{Estimated second-level effects for measure lemmas in GLMMs for Masculine\slash Neuter and Feminine}
%\label{fig:glmmfixef}
%\end{figure}
%
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.6\textwidth]{figures/corpus/04_glmm_fixeff_mn_Measureclass}
%\caption{Effect plot for the masculine\slash neuter GLMM: Measure lemma class effect}
%\label{fig:glmm:fixef:measureclass}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Attraction}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Attraction}
%\caption{Effect plot for the two GLMMs: attraction effect (centered and standardised) from the NAC construction with a bare kind noun and the PGC with a determiner; scales are not aligned between left and right panel}
%\label{fig:glmm:fixef:attraction}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Matchlength}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Matchlength}
%\caption{Effect plot for the two GLMMs: effect of the (centered and standardised) length of the whole construction; scales are not aligned between left and right panel}
%\label{fig:glmm:fixef:matchlength}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Measurecase}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Measurecase}
%\caption{Effect plot for the two GLMMs: effects of the case of the head (kind) noun; scales are not aligned between left and right panel}
%\label{fig:glmm:fixef:measureclass}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Measurenumber}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Measurenumber}
%\caption{Effect plot for the two GLMMs: effect of the grammatical number of the head (kind) noun; scales are not aligned between left and right panel}
%\label{fig:glmm:fixef:measurenumber}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Minus1pos}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Minus1pos}
%\caption{Effect plot for the two GLMMs: effect of part-of-speech preceding the construction; C = cardinal, D = determiner, O = \textit{other}, A = adjective, P = preposition; scales are not aligned between left and right panel}
%\label{fig:glmm:fixef:minus1pos}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.5\textwidth]{figures/corpus/04_glmm_raneff_mn}~\hspace{0.05\textwidth}~\includegraphics[width=0.5\textwidth]{figures/corpus/04_glmm_raneff_fem}
%\caption{Sample random effects for the masculine\slash neuter and feminine GLMMs}
%\label{fig:glmm:fixef:minus1pos}
%\end{figure}

@@

\subsection{Summary and conclusions}
\label{sec:modelssummary}

@@

Concretely, I sampled 16 attested sentences from the corpus and simplified them (mostly removing embedded clauses, complex adverbials, etc.) and standardized their orthography.
It was made sure that the simplifications and normalisations did not affect any of the regressors used in the corpus study, of course.
Eight sentences contained masculine or neuter kind nouns, the other eight contained feminine kind nouns.
Furthermore, in each of the M\slash N and F groups, four sentences with a case drop (\ie\ genitive or oblique case on the kind noun) and four withe case identity were chosen.
More precisely, the four sentences were sampled as highly prototypical examples of case drop (high logits) and case identity (low logits), respectively, according to the corpus based models.%
\footnote{Sentences were sorted by their corresponding $\hat{y}$ calculated from the model.
For the M\slash N sentences, sentences were sampled from the top and bottom 15\% of the resulting vector.
For the F sentences, the top and bottom 25\% were used because the overall sample size was smaller.}
It was made sure that the highly predictive features according to Section \ref{sec:modelssummary} occurred in unique combinations among the four sentences in each cell.
This is summarised in Table \ref{tab:experiment1:design}.

>>>>

%\subsection{Tools for modelling alternations}
%\label{sec:righttools}
%
%In this section, I discuss the relevant statistical approaches to modelling alternations based on corpus data.
%I mainly focus on well-established Generalised Linear Models (GLMs) and their \textit{multi-level} or \textit{hierarchical} extensions, which are also called \textit{mixed models} (GLMMs).
%
%% TODO CONTENT HERE
%
%In Section~\ref{sec:corpusnonhierarchicalmodel}, I briefly discuss why a non-hierarchical model -- while it is an option in principle -- has serious disadvantages for the data set under discussion (and similar data sets).
%In Section~\ref{sec:corpushierarchicalmodel}, I present the hierarchical model which is actually interpreted as the primary result of this study.
%Section~\ref{sec:bayesian} takes a short detour and shows (by example) that results from Bayesian estimation are not intrinsically different from those obtained from Maximum Likelihood methods, although Bayesian approaches have recently been touted to solve various problems facing applied statistics in various fields, even including cognitively oriented corpus linguistics.
%Finally, I summarise and interpret the results of the corpus study in Section~\ref{sec:interpretation}.

@@

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always (but not exclusively) occur in the \PGCa\ and virtually never in the \NACa\ (see Section~\ref{sec:descriptive}).
More indirectly, however, plurality of the measure noun could also favour the \PGCa\ over the \NACa, and exact quantification in the form of cardinals even more so.%
\footnote{This preference was observed empirically by \cite{Hentschel1993}.}
I provide a highly illustrative example using a measure noun which has virtually developed two variants, one of which grammaticalised to a much higher degree, namely \textit{Haufen} `heap\slash lot'.
The phrase \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (\NACa) has essentially two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former use, however, is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps of French cotton' (\PGCa) -- not `two lots of French cotton'.
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of discrete portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.
The statistical models in Section~\ref{sec:corpusstudies} will be designed to detect such a subtle effect, if it exists.
%\footnote{In the extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
%Also note that the presentation of the examples in NAC and PGC is suggestive.
%Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
%However, it was made clear from the beginning and throughout that the preferences are indeed subtle.}

@@

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are positively correlated with the relative frequency the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the NAC.
  \item Plural measure nouns and measure nouns quantified by cardinals favour the PGC variant.
  \item The NAC is associated more strongly with higher registers.
\end{enumerate}

@@

\subsection{Variables and annotation}
\label{sec:annotation}

@@

To capture the influence of register or style mentioned in Section~\ref{sec:analyses}, several proxy variables were used.
At the document level, the DECOW corpus has an annotation for \textit{Badness}.
As described in \cite{SchaeferEa2013}, \textit{Badness} measures how well the distribution of highly frequent short words in the document matches a pre-generated language model for German.
Documents with higher Badness usually contain more incoherent language, shorter sentences, etc.
If the PGC actually favours higher registers and styles, a high \textit{Badness} should be correlated with fewer occurrences of the PGC.
Documents in DECOW14 have also been annotated with a variable called \textit{Genitives}.
The higher the values of this variable, the lower the proportion of genitives among all case-bearing forms is.
While more genitives are also indicative of higher registers, the use of this variable as a regressor in the present study might be considered problematic.
Since the PGC contains a genitive itself, the regressor variable \textit{Genitive} and the document-level variable \textit{Genitives} are not fully independent.
However, since the PGCs make up for only a minute fraction of all genitives, I still use \textit{Genitives} as a regressor with the appropriate caveats.
Since word length is also a good indicator of higher registers and styles, the length of the match in characters (for example 17 for \textit{Kilo tolle Schoki} `a kilo of nice choc' and 49 for \textit{ein Kilogramm vorzüglicher Edelbitter-Schokolade} ` a kilogram of excellent fine dark chocolate') was added as another proxy to register.%
\footnote{The new version of the corpus contains over 100 document-level variables in the spirit of \cite{Biber1988}.
Unfortunately, these annotations were not available when my samples were taken, and even cruder proxies to register than Biberian features (like the length of the query match) had to be used.
All in all, all variables used as indicators of register here have to be interpreted with extreme caution.
}
Furthermore, abbreviated measure nouns like \textit{kg} or \textit{m} might be indicative of less elaborate technical styles, and hence the variable \textit{Measureabbreviated} was included, which is 1 if the measure noun is abbreviated.

The variables \textit{Measurecase} and \textit{Measurenumber} simply encode the grammatical case and number of the measure noun.
The variable \textit{Minus1pos} encodes the part-of-speech of the word preceding the measure noun.
The level names stand for \textit{adjective}, \textit{cardinal}, \textit{determiner}, \textit{other}, and \textit{preposition} (in the same order as in Table~\ref{tab:variables}).
The purpose of this variable is mainly to see whether cardinals really favour the \PGCa\ as hypothesised in Section~\ref{sec:analyses}.

@@

Finally, two variables were added as more or less nuisance variables in the context of the present study.
First, it was reported in the literature that MNPs in the dative and with a masculine or neuter kind noun favour the \PGCa\ more than the corresponding nominative and accusative MNPs \citep{Hentschel1993,Zimmer2015}.
As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (\PGCa) would be preferred over \textit{mit einem Stück frischem Brot} (\NACa).
As with all the examples, native speakers of German will most likely notice that differences are subtle.
To control for this variable, the case of the measure noun was manually annotated (variable \textit{Measurecase}).
Second, to control for lemma frequency effects, the logarithm-transformed frequency per 1,000,000 words of each lemma was added from the frequency lists distributed by the DECOW designers (variables \textit{Kindfreq}, \textit{Measurefreq}).
All numeric variables were standardised and centered (z-transformed) to facilitate their interpretation in the regression models reported 
in the next section.

@@

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation, and \textit{rstanarm} \citep{rstanarm} for Markov-Chain Monte Carlo estimation.
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that either the \NACa\ or the \PGCa\ is chosen.
As explained in Section~\ref{sec:descriptive}, due to extensive case syncrestism in feminine NPs, feminine kind nouns in MNPs like \textit{(mit) einem Liter kalter Milch} `with a litre of cold milk' can either be datives (case agreement, \NACa) or genitives (\PGCa) whenever the measure noun is in the dative.
All dative MNPs with feminine kind nouns had to be removed, and the regressor \textit{Measurecase} consequently only had two levels with feminine kind nouns.%
\footnote{Together with other cleanups (erroneous hits, extremely noisy material), the removal of such uninformative sentences reduced the total size of the sample to 5,063 sentences.}
Therefore, two separate models for masculine\slash neuter and feminine kind nouns were fitted.
For the masculine\slash neuter model, the final sample size was $n_{\text{MN}}=3,949$ with 672 cases of \PGCa\ and 3,277 cases of \NACa.
For the feminine model, it was $n_{\text{F}}=1,114$ with 462 cases of \PGCa\ and 652 cases of $\NACa$.
All regressors from Table~\ref{tab:variables} were included to begin with, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The regressors listed in Table~\ref{tab:variables} with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \cite[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
The results of the estimation are shown in Table~\ref{tab:bigtable}.

@@

Regressors were removed if no convergence could be achieved (even after tuning the optimizer) or perfect separation occurred.
Beyond that, no step-down model selection was performed (see Section~\ref{sec:rightstatistics}).
Generalized variance inflation factors for the remaining regressors were calculated to check for multicollinearity \citep{FoxMonette1992}.
Except for the levels of lemma noun class factors \textit{Kindconsistency} and \textit{Measureclass}, virtually no regressor has a GVIF over 3.
Exceptions were GVIF(\textit{Kindattraction})=3.17 and GVIF(\textit{Kindfreq})=3.92, both still being far from dramatic.
The high GVIFs for the noun lemma classes (up to 46.2) correlate with a low degree of theoretical motivation (at least for the concrete classes as annotated).
However, re-estimating the models with these variables removed did \textit{not} alter the sign or magnitude of any of the other regressors, and I left them in the model.

@@

Standard diagnostics show that the models perform very well.
For the masculine\slash neuter model, Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.495$ and $R^2_c=0.544$ (see \citealp{Gries2015} for a linguist-friendly introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.872$, which means a proportional reduction of error of $\lambda=0.247$.
For the feminine model, $R_m^2=0.468$ and $R^2_c=0.574$ are achieved, as well as a correct prediction rate of $0.815$, which means a proportional reduction of error of $\lambda=0.554$.
The measure and kind lemma intercepts are distributed with the following standard deviations:
Masculine\slash neuter $\sigma_{\text{Measurelemma}}>0.001$ (131 lemmas), $\sigma_{\text{Kindlemma}}>0.591$ (68 lemmas);
Feminine $\sigma_{\text{Measurelemma}}=0.905$ (73 lemmas), $\sigma_{\text{Kindlemma}}=0.000$ (19 lemmas).
While in the masculine\slash neuter model individual measure lemmas do not play much of role, kind lemmas (and -- as will be shown -- the second level model for kind lemmas) are irrelevant in the feminine model, most likely because of the low number of levels.%
\footnote{This low number is expected given that most mass nouns are masculine and neuter in German.}
This will be discussed in Section~\ref{sec:interpretation}.

In Table~\ref{tab:bigtable}, there are groups of columns for the results obtained from Maximum Likelihood estimation and Markov-Chain Monte Carlo estimation, both for the masculine\slash neuter and the feminine models and including confidence intervals.
In the remainder of this section, I describe how these results were obtained.
The coefficient estimates are specified for each regressor (level) in the columns labelled \textit{Coef}.%
\footnote{Empty cells in the table correspond to regressors and their levels which had to be removed because of convergence problems.}
For a robust quantification of the precision of the estimation, I ran a parametric bootstrap (using the \textit{confint.merMod} function from \textit{lme4}) with 1,000 replications, and using the percentile method for the calculation of the intervals.
The resulting 95\% bootstrap confidence intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{Low}, \textit{High} (for the upper and lower 2.5th percentile) and \textit{Width} (for the width of the interval).
The column \textit{Not 0} shows an asterisk for those intervals that do not include 0 as a more modest replacement of significance levels.

As mentioned above, no step-down model selection was performed.
However, for each regressor, a $p$ value was obtained by dropping the regressor from the full model, re-estimating the nested model and comparing it to the full model.
Instead of inexact Wald approximations or Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test from the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}.
These bootstrapped $p$ values are given in the columns labelled p\Sub{BP} in Table~\ref{tab:bigtable}.

Finally, I show that Bayesian methods in the form of Markov-Chain Monte Carlo estimation do not necessarily (and in this case predictably) lead to different results.
Both the masculine\slash neuter model and the feminine model were re-estimated using the \textit{stan\_glmer} function from the \textit{rstanarm} package, which provides an \textit{lme4}-compatible syntax for estimating common model types with the \textit{stan} software \citep{CarpenterEa2017}.
The algorithm was run using 4 chains and 1,000 iterations, and instead of guessing prior distributions or using implausible uniform prior distributions \cite[251--252]{Levshina2016}, I used plausible default priors.
Most notably, priors for coefficients were specified as $\mathcal{N}(0,10)$ because coefficients higher than $10$ or lower than $-10$ are extremely rare in well-specified models on appropriate data.%
\footnote{Consider that with a coefficient of 10, each increase by $1$ in the regressor variable means an increase in odds of $exp(10)=22026.47$.
To reliably estimate such coefficients, extremely large samples would be required.}
The algorithm converged, and for all coefficients, the $\hat{R}$ diagnostic was exactly $1$.
The resulting coefficients and intervals as well as an * for \textit{interval does not contain $0$} are also given in Table~\ref{tab:bigtable} in the groups of columns labelled \textit{MCMC}.
In the interpretation of the results in the next section, it will become apparent that both methods of estimation lead to the exact same theoretically relevant results.

@@

%\begin{figure}[h]
%\centering
%\includegraphics[width=0.8\textwidth]{figures/corpus/04_glmm_fixef_firstlevel}
%\caption{Estimated first-level effects in GLMMs for Masculine\slash Neuter and Feminine}
%\label{fig:glmmfixef}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.8\textwidth]{figures/corpus/04_glmm_fixef_secondlevel_kind}
%\caption{Estimated second-level effects for kind lemmas in GLMMs for Masculine\slash Neuter and Feminine}
%\label{fig:glmmfixef}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.8\textwidth]{figures/corpus/04_glmm_fixef_secondlevel_measure}
%\caption{Estimated second-level effects for measure lemmas in GLMMs for Masculine\slash Neuter and Feminine}
%\label{fig:glmmfixef}
%\end{figure}
%

\subsection{Interpretation}
\label{sec:interpretation}

The results reported in Section~\ref{sec:corpushierarchicalmodel} generally confirm the hypotheses from Section~\ref{sec:analyses}.
Before I turn to some details, it should be noticed that the results from standard MLE (with bootstrap) and MCMC (with default priors) obviously differ numerically but not enough to result in different interpretations.
The signs and magnitudes of the coefficients are identical with the two methods, and the confidence intervals have roughly the same width and symmetry properties.
In only one single case (masculine\slash neuter, \textit{Measureattraction}) the MLE bootstrap confidence interval does not include $0$ while the MCMC interval does (barely, with $-0.01$).
While I only interpret the MLE results, virtually the same interpretation could be made for the MCMC results.
Furthermore, the masculine\slash neuter model and the feminine model agree in many relevant aspects.
Disagreement between the models will be discussed below.

% ATTRACTION

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Measureattraction}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Measureattraction}
  \caption{Effect plot for \textit{Measureattraction}}
  \label{fig:eff:measureattraction}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_mn_Kindattraction}~\includegraphics[width=0.4\textwidth]{figures/corpus/04_glmm_fixeff_fem_Kindattraction}
  \caption{Effect plot for \textit{Kindattraction}}
  \label{fig:eff:kindattraction}
\end{figure}

First, the prototypicality effect related to the non-alternating \PGCd\ and \NACb\ can be shown, see the effect plots in Figure~\ref{fig:eff:measureattraction} and Figure~\ref{fig:eff:kindattraction}.%
\footnote{Effect plots were created using the \textit{effects} package \citep{Fox2003}.
It shows the change in probability for the outcome (y axis) dependent on values of a regressor (x axis),assuming typical values of all other regressors.}
The effect is mostly as expected:
if a lemma appears relatively more often in the \PGCd\ (compared to its frequency in the \NACb), the more often the \PGCa\ is chosen over the \NACa\ with this specific lemma.
Only the feminine model \textit{Kindattraction} regressor is clearly useless (`non-significant') in the PBmodcomp test, and the effect plot shows a corresponding picture.
This is related to the fact that in the feminine model the kind lemma effect and its second-level model turned out to be uninformative relative to the alternation (remember that $\sigma_{\text{Kindlemma}}=0.000$).

% MN
% $\sigma_{\text{Measurelemma}}<0.001$
% $\sigma_{\text{Kindlemma}}>0.591$
% F
% $\sigma_{\text{Measurelemma}}=0.905$
% $\sigma_{\text{Kindlemma}}=0.000$

% GRAMMATICALISATION
% => lemmas and lemma classes

% PLURAL & CARDINALS

% REGISTER
% => Badness
% => Genitives
% => Matchlength
% => Measureabbreviated

% NUISANCE
% => dative effect
% => frequency

@@

Concretely, I sampled 16 attested sentences from the corpus and simplified them (mostly removing embedded clauses, complex adverbials, etc.) and standardized their orthography.
It was made sure that the simplifications and normalisations did not affect any of the regressors used in the corpus study, of course.
Eight sentences contained masculine or neuter kind nouns, the other eight contained feminine kind nouns.
Furthermore, in each of the M\slash N and F groups, four sentences with a case drop (\ie\ genitive or oblique case on the kind noun) and four withe case identity were chosen.
More precisely, the four sentences were sampled as highly prototypical examples of case drop (high logits) and case identity (low logits), respectively, according to the corpus based models.%
\footnote{Sentences were sorted by their corresponding $\hat{y}$ calculated from the model.
For the M\slash N sentences, sentences were sampled from the top and bottom 15\% of the resulting vector.
For the F sentences, the top and bottom 25\% were used because the overall sample size was smaller.}
It was made sure that the highly predictive features according to Section \ref{sec:interpretation} occurred in unique combinations among the four sentences in each cell.
This is summarised in Table \ref{tab:experiment1:design}.
